---
title: "Lab 02  RMD, more ggplot (Bella Fascendini)"
author: "Bella Fascendini"
date: "09/18/2024"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
setwd("/Users/bf2383/Downloads")
#knitr::purl("PSY503_lab2.Rmd")
```

# PSY 503: Lab 02: R markdown, and ggplot

**09/18/2024**

For Lab 2, please carry out the following tasks. Make sure to install/load any relevant libraries not explictly specified below. At the end of the lab, please submit the final rmd file (important) along with other files you generated (QMD, PDF,HTML, docx, and r)


## Setup Tasks

0.  install the package tinytex into your system using the following command.

```{r, eval=FALSE}
install.packages('tinytex') 
tinytex::install_tinytex() 
# to uninstall TinyTeX, run tinytex::uninstall_tinytex()
```

```{r}
library(tidyverse)
```

Tinytex, is needed for creating pdf reports via R-markdown.

## Task with YAML Header (A)

-   Edit the YAML header / metadata to reflect your name.

## Task with Markdown and R chunks

-   During Monday's class, we learned that the sum of squared deviations from mean for a given set of data SS is given by $$SS = \sum_{i=1}^N (\bar{X} - x_i)^2$$ Knowing this, what is the full formula for variance? Type below in markdown, and make sure it is in Bold. (Hint: Variance when we are working with a population = $SS/N$, and Variance of a sample is $SS/N-1$)

$$\text{Variance} = \frac{\mathbf{SS}}{\mathbf{N-1}}$$

```{r}
compute_SS <- function(x) {
  n <- length(x)
  mean_x <- mean(x)
  sum_squared_diff <- sum((x - mean_x)^2)
  return(sum_squared_diff)
}
vector_a <- c(5, 5, 5, 5, 5)
vector_b <- c(1, 3, 5, 7, 9)

```

## Descriptives (dplyr) and Visualization tasks

In the previous lab, we worked with a data from DataSaurus. We visualized y against x for different values of the column named *"dataset"*. Now, carry out this visualization but with the following changes.

a.  show all the different visualizations within a single graph using *facets* in ggplot
b.  make sure that each dataset is colored differently
c.  great, now a journal has accepted your work for publication. but they charge \$10,000 for colored graphs. to save your lab some resources, redo the graph so that the colors are different shades of grey. (Hint: look at *scale_colour_grey*)
d.  you realize that given so many different conditions/datasets, it's a bit harder to map the color to condition. to make it easy for the reader and to be robust against bad print quality, change the shape of the plotted points based on the condition.
e. out of curiousity, you also create a new plot in a different direction: this doesn't plot data points for each condition but just creates a boxplot for each dataset (no need to facet). what does this plot look like?

```{r plots}
library(datasauRus)
library(MetBrewer)
lab2_datasaurus<- datasaurus_dozen

# Produce graphs for the different changes below

#a r plot using facets in ggplot
ggplot(lab2_datasaurus, aes(x = x, y = y)) +
  geom_point() +
  facet_wrap(~ dataset) +
  theme_minimal()

#b plotting each dataset in different color
ggplot(lab2_datasaurus, aes(x = x, y = y, color = dataset)) +
  geom_point() +
  facet_wrap(~ dataset) +
  theme_minimal() +
  scale_color_met_d(name = "Archambault")

#c plotting in different shades of grey
ggplot(lab2_datasaurus, aes(x = x, y = y, color = dataset)) +
  geom_point() +
  facet_wrap(~ dataset) +
  theme_minimal() +
  scale_colour_grey(start = 0.2, end = 0.8)

#d changing shape of plotted points
ggplot(lab2_datasaurus, aes(x = x, y = y, color = dataset, shape = dataset)) +
  scale_shape_manual(values = 1:13) +
  geom_point() +
  facet_wrap(~ dataset) +
  theme_minimal() +
  scale_colour_grey(start = 0.2, end = 0.8)

#e box plot
ggplot(lab2_datasaurus, aes(x = dataset, y = y)) +
  geom_boxplot() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1.5)) 

```


You've learned that looking at the data is important. But summary statistics can be useful too. The following dplyr-based code computes the mean of `lab2_datasaurus` for each dataset within it, for the variable x and stores it into a new dataframe. 

```{r}
mean_datasaurus <- lab2_datasaurus %>% 
    group_by(dataset) %>% 
    summarize(
      mean_x    = mean(x)
    )

print (mean_datasaurus)
```
- produce a dataframe that has not just mean of x, but also the following summary statistics in appropriately named columns:
-- mean of y
-- median of x
-- median of y
-- variance of x
-- variance of y
-- correlation between x and y (found using the *cor()*)

What do you observe?

Despite having similar summary statistics (mean, median, variance, etc), each dataset looks VERY different visually. This is a good reminder to not solely rely on summary statisctics when anlayzing datasets.

```{r}
summary_datasaurus <- lab2_datasaurus %>%
  group_by(dataset) %>%
  summarize(
    mean_y = mean(y),
    median_x = median(x),
    median_y = median(y),
    variance_x = var(x),
    variance_y = var(y),
    correlation_xy = cor(x, y)
  )
print(summary_datasaurus)
```


## Combining two sources of data in a graph

Now let's try overlaying the (mean of x, mean of y) point for the different datasets on this graph. To distinguish it from the other datapoints you can make this overlaid plot larger in size.  

```{r}
# calculating the mean of x and y
mean_values <- lab2_datasaurus %>%
  group_by(dataset) %>%
  summarize(
    mean_x = mean(x),
    mean_y = mean(y)
  )
ggplot(lab2_datasaurus, aes(x = x, y = y)) +
  geom_point() + # Scatter plot of the data points
  geom_point(data = mean_values, aes(x = mean_x, y = mean_y), color = "magenta", size = 5) + 
  facet_wrap(~ dataset) +
  theme_minimal() +
  labs(
    x = "X values",
    y = "Y values"
  )
```

## RMD vs. Quarto
Open a new Quarto document. Attempt translating from this document into Quarto the first 3 markdown and code chunks of this document. What differences do you notice?

Not a lot. I think visually it looks a bit more appelaing in Quarto because all the text is displayed already formatted compared to R markdown. I'd need to do an extra step of knitting to html/pdf for the text to be formatted.

## Tasks with YAML Header (B)
You know that the output type is indicated in the YAML header. By changing the header or using the knit dropdown, create outputs in Pdf, html, and docx. 


## Extracting the code back to r

On some occasions you have an r markdown file, but you may want to get ride of the markdown and yaml, and extract just the r-code. This is possible by usking the following command 

*knitr::purl("filename.Rmd")*
```{r}
knitr::purl("/Users/bf2383/Downloads/PSY503_lab2.Rmd", output = "PSY503_lab2.R")
```

As an exercise, use this command to save your current rmd file as an R file. 

## End of lab.