---
title: "Lab 3- Data Wrangling - Questions"
author: "Bella Fascendini"
date: "2024-09-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gapminder)
```
# Lab 3 - Gapminder Data Wrangling Lab Assignment

Using the Gapminder dataset, complete the following tasks. Use tidyverse and dplyr functions and pipes where appropriate.
Remember to load the necessary libraries and the Gapminder dataset before starting.

1) Find all countries in Asia with a life expectancy greater than 75 years in 2007.
```{r}
head(gapminder)

asia_75plus_2007 <- gapminder %>%
  filter(year == 2007, continent == "Asia", lifeExp > 75) %>%
  select(country, lifeExp)

asia_75plus_2007

```

2) List the top 5 countries with the highest GDP per capita in 2007, in descending order.
```{r}
top_5_gdp <- gapminder %>%
  filter(year == 2007) %>%
  arrange(desc(gdpPercap)) %>%
  select(country, gdpPercap) %>%
  slice(1:5)
top_5_gdp
```

3) Create a new dataframe with only the country, continent, and GDP per capita for the year 2007.
```{r}
df_2007 <- gapminder %>%
  filter(year == 2007) %>%
  select(country, continent, gdpPercap)

df_2007
```

4) Add a new column called "gdp" that calculates the total GDP (GDP per capita * population) for each country in 2007.\
```{r}
gdp_total <- df_2007 %>%
  mutate(gdp = gdpPercap * gapminder$pop[gapminder$year == 2007])

gdp_total
```

5) Create two separate dataframes: one for countries in Europe and one for countries in Africa, both for the year 2007. Then bind these dataframes together.
```{r}
europe_2007 <- gapminder %>%
  filter(year == 2007, continent == "Europe")

africa_2007 <- gapminder %>%
  filter(year == 2007, continent == "Africa")

europe_africa_combined <- bind_rows(europe_2007, africa_2007)

europe_africa_combined
```

6) Calculate the average life expectancy and total population for each continent in 2007.
```{r}
summary_2007 <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarize(
    average_life_expectancy = mean(lifeExp),
    total_population = sum(pop)
  )

summary_2007

```

7) Create a wide format dataframe that shows the population for each country across all life expectancies as columns.

```{r}
population_wide <- gapminder %>%
  #filter(year == 2007) %>%
  select(country, pop, year) %>%
  pivot_wider(names_from = country, 
              values_from = pop,
              names_prefix = "pop_")
population_wide
```
